|

### 2021-08-10 prova_knit_finale.Rmd
altre modifiche

Let us consider a more general state-space model, where the state vector includes a time-constant parameter vector $\psi \in \Psi$, where $\Psi$ is the parameter space. Since we can interpret the parameter as a state with the law of motion $\psi_t=\psi_{t-1}=\psi$, then it is possible to adopt a PF algorithm to estimate $\psi$.\
However, being $\psi$ constant over time, it is meaningless to sample its values sequentially: the first extraction of $\psi$ at the initial stage 0 (i.e., $\psi_0^n$) generates a constant path of particles for the parameter ($\psi_t^n=\psi_0^n$). \footnote{To be more precise, at the resampling step, also $\psi^n$ is resampled, but the possible values that it may take are the $N$ ones obtained at the initial time 0 extraction.}\
Therefore, Liu and West (2001) proposed a modified particle filter, called \textit{Liu and West filter} (LWF) that allows to resample the parameter over time from a continuous distribution. In this way, at every time, the support of $\psi$ is not limited to the $N$ initially sampled values. \
Here, we describe the LWF that makes use of the normal distribution to sample sequentially the parameter. Moreover, consider the simpler version of the bootstrap filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF. For instance, in our following implementation with the random walk model we use an auxiliary particle LWF.}\
Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi(\cdot)$ be the prior distribution for the parameter $\psi$.\
The LWF can be described by the following algorithm.
\begin{enumerate}
    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\Pd_0(dx_0)$ and $\psi_0^n\thicksim \pi(\psi_0)$.

@@

    \item Set the indexes: $A_t^{n}=n$ for  $n=1,...,N$

@@

At the initial step, $N$ states and $N$ parameter values are sampled independently, respectively, from $\Pd_0$ and $\pi$ and the particle weights are set to be constant, as usual.\

@@

        Var_{\pi_{t-1}^N} ( \psi)=&Var\big[\E( \psi|A_t^{n})\big]+\E \big[Var( \psi|A_t^{n})\big]=Var\big[\psi^{A_t^{n}}\big]+\E\big[\Omega\big]=2\cdot\Omega

@@

    \pi^N_{t}(\psi)=\pi^N(\psi|y_{1:t})=\sum_{n=1}^NW_{t}^n\delta_{\psi^n}(\psi)

@@

\subsection{Implementation}
\
Consider the linear Gaussian example of section \ref{implementation}, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$. Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
\begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*}
Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. Note that in this model we sample from a Gamma distribution (instead of the Normal one, as in the previous description of the LWF) since the parameters of interest are variances, that can take only positive values. Next, we descrive the algorithm.
\begin{algorithm} LWF for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$. Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$. Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
\item For $t=1,...,n$:.
\begin{enumerate}
\item Compute
\begin{align*}
\overline{\sigma^{2}}& =E_{\hat{\pi}_{t-1}}(\sigma^{2})\\
\overline{\tau^{2}}& =E_{\hat{\pi}_{t-1}}(\tau^{2}) \\
\Sigma_{v} & =V_{\hat{\pi}_{t-1}}(\sigma^{2}) \\
\Sigma_{w} & =V_{\hat{\pi}_{t-1}}(\tau^{2})
\end{align*}
then set
\begin{align*}
m_{v}^{(i)} & = a{\sigma^{2}}^{(i)}+(1-a)\overline{\sigma^{2}} \\
{s_{v}^{2}}^{(i)} & =(1-a^{2})\Sigma_{\sigma^{2}} \\
\alpha_{v}^{(i)} & =\frac{(m_{v}^{(i)})^{2}}{{s_{v}^{2}}^{(i)}} \\
\beta_{v}^{(i)} & =\frac{(m_{v}^{(i)})}{{s_{v}^{2}}^{(i)}}\\
m_{w}^{(i)} & = a{\tau^{2}}^{(i)}+(1-a)\overline{\tau^{2}} \\
{s_{w}^{2}}^{(i)} & =(1-a^{2})\Sigma_{\tau^{2}}\\
\alpha_{w}^{(i)} & =\frac{(m_{w}^{(i)})^{2}}{{s_{w}^{2}}^{(i)}} \\
\beta_{w}^{(i)} & =\frac{(m_{w}^{(i)})}{{s_{w}^{2}}^{(i)}}
\end{align*}
\item For $k=1,...,N$:
\begin{itemize}
\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),m_{v}) $
\item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
\item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},{\tau^2}^{(k)})$
\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},{\sigma^2}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},m_{v}^{(I_k)}))}$
\end{itemize}
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{LWfun} function goes through the illustrated steps.
\
\hrule
\hrule
\
\code{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}
\
\hrule
\textbf{Arguments}

@@

\section{Convergence}\label{pf_converg}
In this section, we cover three propositions that justify the use of the PF algorithms to estimate quantities of interest. Firstly, we present two convergence results for the estimators of the predictive and filtering distributions. As we will discuss, these results hold under reasonable conditions. Then, we will show that the error of the particle estimators are asymptotically distributed according to a Gaussian distribution with known variance. 
Henceforth, we will maintain the assumption that (multinomial) resampling is repeated at each iteration.
\subsection{MSE convergence}
As already seen in subsection \ref{seq_impt_sampling}, the particle approximation of $\Q_{t}(dx_{0:t})$ is:
\begin{equation*}
      \sum_{n=1}^NW_{t-1}^nM_t(X_{t-1}^n,dx_t).
\end{equation*}
The following lemma states that the (mean squared) error related to sampling from this conditional distribution is bounded.
\bigskip
\begin{lemma}
For any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
$$
\E\Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg\}^2 \Bigg]\leq \frac{\|\varphi\|_{\infty}^2}{N},
$$
where $\|\cdot\|_{\infty}$ is the supnorm and $M_t\varphi$ is the map 
$x_{t-1}\mapsto \int_{\X}M_t(x_{t-1},dx_t)\varphi(x_t).$
\end{lemma}
This result can be used to prove the convergence of the estimator $\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)$. To see this, note that 
\begin{equation}
    \begin{split}
        \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)=\bigg(&\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg) \\
        & + \bigg(\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n)- \E_{\Q_{t-1}M_t}(\varphi)\bigg).
    \end{split}
\end{equation}
where $\E_{\Q_{t-1}M_t}(\varphi)=\int_{\X^{t+1}}\varphi(x_t)  \Q_{t-1}(dx_{0:t-1})M_t(x_{t-1},dx_t)$.\
By the lemma, the first component of the right-hand side has a bounded MSE. It can be proved that also the second component of the right-hand side has a bounded MSE. Using the fact that $\E\{(X+Y)^2\}\leq2\E\{\E(X^2)+\E(Y^2)\},$ it follows that also the MSE of the left-hand side is bounded. A simple induction argument proves that the result holds for all $t$.\
Using a similar argument, one can prove that the MSE of the estimator $\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)$ is bounded as well. These ideas are formalized in the following proposition.
\bigskip
\begin{proposition}
For any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$ $\forall t\geq0$ there exist constants $c_t,c_t'$ such that:
$$
\E \Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)\bigg\}^2 \Bigg]\leq c_t \frac{\|\varphi\|_{\infty}^2}{N}
$$
and 
$$
    \E \Bigg[ \bigg\{\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\E_{\Q_t}(\varphi) \bigg\}^2 \Bigg]\leq c_t' \frac{\|\varphi\|_{\infty}^2}{N}.
$$
Therefore, taking the limit for $N\rightarrow \infty,$ we have that 
$$
\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\xrightarrow{\text{MSE}}\E_{\Q_{t-1}M_t}(\varphi)
$$
and 
$$
\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\xrightarrow{\text{MSE}}\E_{\Q_t}(\varphi)
$$
\end{proposition}
In the next subsection, we establish a much stronger property of the particle filter algorithms.

>>>>

Let us consider a more general state-space model, where the state vector includes a time-constant parameter vector $\psi \in \Psi$, where $\Psi$ is the parameter space. Since we can interpret the parameter as a state with the law of motion $\psi_t=\psi_{t-1}=\psi$, then it is possible to adopt a PF algorithm to estimate $\psi$.\
However, being $\psi$ constant over time, it is meaningless to sample its values sequentially: the first extraction of $\psi$ at the initial stage 0 (i.e., $\psi_0^n$) generates a constant path of particles for the parameter ($\psi_t^n=\psi_0^n$). \footnote{To be more precise, at the resampling step, also $\psi^n$ is resampled, but the possible values that it may take are the $N$ ones obtained at the initial time 0 extraction.}\
Therefore, Liu and West (2001) proposed a modified particle filter, called \textit{Liu and West filter} (LWF) that allows to resample the parameter over time from a continuous distribution. In this way, at every time, the support of $\psi$ is not limited to the $N$ initially sampled values. \
Here, we describe the LWF that makes use of the normal distribution to sample sequentially the parameter. Moreover, consider the simpler version of the bootstrap filter.\footnote{Clearly, the LWF can be also used within the more general frameweorks of the GPF and the APF. For instance, in our following implementation with the random walk model we use an auxiliary particle LWF.}\
Let the transition kernel and the likelihood depend on the parameter $\psi$ (i.e., $P_t(x_{t-1},dx_t;\psi)$ and $f_t(y_t|x_t;\psi)$) and let $\pi_0(\cdot)$ be the prior distribution for the parameter $\psi$.\
The LWF can be described by the following algorithm.
\begin{enumerate}
    \item \textbf{Inital stage: }For $n=1,...,N$, at time 0 sample: $X_0^n\thicksim\Pd_0(dx_0)$ and $\psi^n\thicksim \pi_0(\psi)$.

@@

@@

At the initial step, $N$ states and $N$ parameter values are sampled independently, respectively, from $\Pd_0$ and $\pi_0$ and the particle weights are set to be constant, as usual.\

@@

        Var_{\pi_{t-1}^N} ( \psi)=&Var\big[\E( \psi|A_t^{n})\big]+\E \big[Var( \psi|A_t^{n})\big]=Var\big[\psi^{A_t^{n}}\big]+\E\big[\Omega\big]>\Omega

@@

    \pi^N_{t}(\psi)=\pi^N(\psi|y_{1:t})=\sum_{n=1}^NW_{t}^n\delta_{\psi^n}(d\psi)

@@

\subsection{Implementation}
Consider the linear Gaussian example of section \ref{implementation}, but this time with unknown variances $\tau^{2}$ and $\sigma^{2}$. Thus, let $\psi=(\sigma^{2},\tau^{2})$ be the unknown parameter vector and assign a gamma prior for its components,
\begin{align*}
\sigma^{2}  & \sim G(\alpha_{v},\beta_{v}) \\
\tau^{2}  & \sim G(\alpha_{w},\beta_{w})
\end{align*}
Alternatively, assign them a uniform prior if we have no knowledge of the hyperparameters. Note that in this model we sample from a Gamma distribution (instead of the Normal one, as in the previous description of the LWF) since the parameters of interest are variances, that can take only positive values. Note that we are applying the LWF in the framework of APF with two resampling stage and the correct transition kernel. Next, we describe the algorithm.
\begin{algorithm} LWF for Random Walk plus Noise Model
\begin{itemize}
\item Initialize $(x_{0}^{(1)},...,x_{0}^{(N)})$ from $N(m_{0},C_{0})$, $({\sigma^{2}}^{(1)},...,{\sigma^{2}}^{(N)})$ from $G(\alpha_{v},\beta_{v})$ and $({\tau^{2}}^{(1)},...,{\tau^{2}}^{(N)})$ from $G(\alpha_{w},\beta_{w})$. Set $w_{0}^{(i)}=N^{-1} \ \forall \ i=1,...,N$. Therefore $\psi^{(i)}=({\sigma^{2}}^{(i)},{\tau^{2}}^{(i)})$, and
$\hat{\pi}_{0}=p(x_{0}|y_{0})=\sum_{i=1}^{N}w_{0}^{(i)}\delta_{(x_{0}^{(i)},\psi^{(i)})}$
\item For $t=1,...,n$:.
\begin{enumerate}
\item Compute
\begin{align*}
\overline{\sigma^{2}}& =E_{\hat{\pi}_{t-1}}(\sigma^{2})\\
\overline{\tau^{2}}& =E_{\hat{\pi}_{t-1}}(\tau^{2}) \\
\Sigma_{v} & =V_{\hat{\pi}_{t-1}}(\sigma^{2}) \\
\Sigma_{w} & =V_{\hat{\pi}_{t-1}}(\tau^{2})
\end{align*}
then set
\begin{align*}
m_{v}^{(i)} & = a{\sigma^{2}}^{(i)}+(1-a)\overline{\sigma^{2}} \\
{s_{v}^{2}}^{(i)} & =(1-a^{2})\Sigma_{\sigma^{2}} \\
\alpha_{v}^{(i)} & =\frac{(m_{v}^{(i)})^{2}}{{s_{v}^{2}}^{(i)}} \\
\beta_{v}^{(i)} & =\frac{(m_{v}^{(i)})}{{s_{v}^{2}}^{(i)}}\\
m_{w}^{(i)} & = a{\tau^{2}}^{(i)}+(1-a)\overline{\tau^{2}} \\
{s_{w}^{2}}^{(i)} & =(1-a^{2})\Sigma_{\tau^{2}}\\
\alpha_{w}^{(i)} & =\frac{(m_{w}^{(i)})^{2}}{{s_{w}^{2}}^{(i)}} \\
\beta_{w}^{(i)} & =\frac{(m_{w}^{(i)})}{{s_{w}^{2}}^{(i)}}
\end{align*}
\item For $k=1,...,N$:
\begin{itemize}
\item Draw $I_{k}$, with $P(I_{k}=i) \propto w_{t-1}^{(i)}f_{N}(y_{t}|g(x_{t-1}^{(i)}),m_{v}) $
\item Draw ${\sigma^{2}}^{(k)}$ from $G(\alpha_{v}^{(I_k)},\beta_{v}^{(I_k)})$ 
\item Draw ${\tau^{2}}^{(k)}$ from $G(\alpha_{w}^{(I_k)},\beta_{w}^{(I_k)})$
\item Draw $x_{t}^{(k)}$ from $N(x_{t-1}^{(I_k)},{\tau^2}^{(k)})$
\item Set  $\tilde{w}_{t}^{(k)} = \frac{f_{N}(y_{t}|x_{t}^{(k)},{\sigma^2}^{(k)})}{f_{N}(y_{t}|g(x_{t-1}^{(I_{k})},m_{v}^{(I_k)}))}$
\end{itemize}
\item Compute $ESS=\Bigg(\sum_{i=1}^{N}(w_{t}^{(i)})^{2}\Bigg)^{-1}$
\item if $ESS<N/2$ then
\begin{enumerate}
\item Draw a sample of size N, $(x_{t}^{(1)},...,x_{t}^{(N)})$, from the discrete distribution $P(x_{t}=x_{t}^{(i)})=w_{t}^{(i)},\ \ i=1,...,N$
\item Reset the weights: $w_{t}^{(i)}=N^{-1}$, $i=1,...,N$.
\end{enumerate}
\item Set $\hat{\pi}_{t}=p(x_{t}|y_{1:t})=\sum_{i=1}^{N}w_{t}^{(i)}\delta_{(x_{t}^{(i)},\psi^{(i)})}$
\end{enumerate}
\end{itemize}
\end{algorithm}
Our \code{LWfun} function goes through the illustrated steps.
\
\hrule
\hrule
\
\code{LWfun(data,N,m0,C0,alphav,betav,alphaw,betaw,delta,unif,r)}
\
\hrule
\textbf{Arguments}

@@

\section{Convergence}\label{pf_converg}
In this section,\footnote{This section has the purpose to give only a very general idea about convergence of PF estimators.} we cover three propositions that justify the use of the PF algorithms to estimate quantities of interest. Firstly, we present two convergence results for the estimators of the predictive and filtering distributions. As we will discuss, these results hold under reasonable conditions. Then, we will show that the error of the particle estimators are asymptotically distributed according to a Gaussian distribution with known variance. All the convergence results that we present here involve the limit for the number of particles $N$ at each iteration.\
Henceforth, we will maintain the assumption that (multinomial) resampling is repeated at each iteration.
\subsection{MSE convergence}
As already seen in subsection \ref{seq_impt_sampling}, the particle approximation of $\Q_{t}(dx_{0:t})$ is:
\begin{equation*}
      \sum_{n=1}^NW_{t-1}^nM_t(X_{t-1}^n,dx_t).
\end{equation*}
The following lemma states that the (mean squared) error related to sampling from this conditional distribution is bounded.
\bigskip
\begin{lemma}
For any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$
$$
\E\Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg\}^2 \Bigg]\leq \frac{\|\varphi\|_{\infty}^2}{N},
$$
where $\|\cdot\|_{\infty}$ is the supnorm and $M_t\varphi$ is the map 
$x_{t-1}\mapsto \int_{\X}M_t(x_{t-1},dx_t)\varphi(x_t).$
\end{lemma}
This result can be used to prove the convergence of the estimator $\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)$. To see this, note that 
\begin{equation}
    \begin{split}
        \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)=\bigg(&\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n) \bigg) \\
        & + \bigg(\sum\limits_{n=1}^NW_{t-1}^n(M_t\varphi)(X_{t-1}^n)- \E_{\Q_{t-1}M_t}(\varphi)\bigg).
    \end{split}
\end{equation}
where $\E_{\Q_{t-1}M_t}(\varphi)=\int_{\X^{t+1}}\varphi(x_t)  \Q_{t-1}(dx_{0:t-1})M_t(x_{t-1},dx_t)$.\
By the lemma, the first component of the right-hand side has a bounded MSE. It can be proved that also the second component of the right-hand side has a bounded MSE. Using the fact that $\E\{(X+Y)^2\}\leq2\E\{\E(X^2)+\E(Y^2)\},$ it follows that also the MSE of the left-hand side is bounded. A simple induction argument proves that the result holds for all $t$.\
Using a similar argument, one can prove that the MSE of the estimator $\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)$ is bounded as well. These ideas are formalized in the following proposition.
\bigskip
\begin{proposition}
For any measurable, integrable, and bounded function $\varphi:\X\rightarrow\R,$ $\forall t\geq0$ there exist constants $c_t,c_t'$ such that:
$$
\E \Bigg[ \bigg\{ \frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)-\E_{\Q_{t-1}M_t}(\varphi)\bigg\}^2 \Bigg]\leq c_t \frac{\|\varphi\|_{\infty}^2}{N}
$$
and 
$$
    \E \Bigg[ \bigg\{\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)-\E_{\Q_t}(\varphi) \bigg\}^2 \Bigg]\leq c_t' \frac{\|\varphi\|_{\infty}^2}{N}.
$$
Therefore, taking the limit for $N\rightarrow \infty,$ we have that 
$$
\frac{1}{N} \sum\limits_{n=1}^N \varphi(X_t^n)\xrightarrow{\text{MSE}}\E_{\Q_{t-1}M_t}(\varphi)
$$
and 
$$
\sum\limits_{n=1}^N W_t^n \varphi(X_t^n)\xrightarrow{\text{MSE}}\E_{\Q_t}(\varphi)
$$
\end{proposition}
In the next subsection, we establish a much stronger property of the particle filter algorithms.
